---
title: "Building the GPW Livestock Census Database"
subtitle: "a workshop on tabshiftr and arealDB"
author: "Steffen Ehrmann · resonantia research gUG"
format:
  revealjs:
    theme: [default, theme.scss]
    slide-number: c/t
    incremental: false
    code-line-numbers: false
    highlight-style: github-dark
    width: 1280
    height: 720
    margin: 0.06
    fig-cap-location: bottom
    transition: none
    pdf-separate-fragments: false
    progress: false
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

```{r setup}
#| include: false
#| eval: true
# Loaded once. tabshiftr's example data is used on the archetype slides.
suppressPackageStartupMessages({
  library(tabshiftr)
  library(knitr)
  library(kableExtra)
})
```


## The integration problem
::: {.subtitle}
areal data come from many sources, none of them are tidy
:::

::: {.row}
::: {style="flex: 1.2"}

19 countries · many decades · different formats

::: {.issue}
[Different file formats]{.what}<br>
[csv, xlsx, shp, gpkg, dbf, pdf]{.why}
:::

::: {.issue}
[Different table layouts]{.what}<br>
[headers in row 1, 2, or 5; multi-row headers; merged cells]{.why}
:::

::: {.issue}
[Different territorial naming]{.what}<br>
["USA", "United States", "U.S.A." — same place, three IDs]{.why}
:::

::: {.issue}
[Different concept vocabularies]{.what}<br>
["cattle", "bovinos", "existencias bovinas" — same animal]{.why}
:::

::: {.issue}
[Sources update]{.what}<br>
[your hand-cleaned spreadsheet breaks every year]{.why}
:::

:::
::: {.box}
[The cost of doing this by hand]{.box-title}

**Error-prone**
Hand-edits hide silent typos and column mis-shifts.

**Unreproducible**
No record of what was changed and why.

**Brittle**
Next year's update breaks last year's pipeline.
:::
:::

## Two packages, one workflow

::: {.row}
::: {.box}
[tabshiftr]{.box-title}
*table reshaping*

Write a **schema** — a formal description of where each variable lives in a raw messy table.

`reorganise()` mechanically applies it to produce a tidy output.

*Same schema, next year's file → same tidy output.*
:::

::: {.box}
[arealDB]{.box-title}
*integration pipeline*

Ingests tables and spatial geometries.

Harmonises territories against a **gazetteer**.

Harmonises concepts against an **ontology**.

Stores everything in a standardised directory.
:::
:::

::: {.footnote .green}
→ tabshiftr makes the table tidy; arealDB makes the database harmonised.
:::


# 1. tabshiftr

::: {.tagline}
schemas for messy tables
:::

## The schema is a description, not a script

A schema describes **WHERE** each variable lives in the raw file.
It does *not* describe **HOW** to reshape — that's `reorganise()`'s job.

<br>

::: {.row}
::: {}
**Raw file**

```{r}
#| eval: true
#| echo: false
df <- data.frame(
  year = c("2008", "2008", "2009"),
  ADM1 = c("Khentii", "Khentii", "Khentii"),
  ADM2 = c("Bayan-Adarga", "Batnorov", "Bayan-Adarga"),
  cattle = c("12450", "9810", "13100"),
  sheep = c("8230", "11200", "8550")
)
kable(df, format = "html") |>
  kable_styling(bootstrap_options = "condensed", font_size = 14)
```
:::

::: {}
**Schema** (says where each variable lives)

```r
setIDVar(name = "year",   columns = 1) |>
setIDVar(name = "ADM1",   columns = 2) |>
setIDVar(name = "ADM2",   columns = 3) |>
setIDVar(name = "animal", columns = c(4, 5),
                          rows    = 1) |>
setObsVar(name = "number_heads", columns = c(4, 5))
```
:::
:::

::: {.footnote .green}
Run `reorganise()` → tidy long table. Same schema works on next year's file.
:::

## Step zero: read the raw file
::: {.subtitle}
as tabshiftr will see it
:::

```r
input <- read_csv(file = "path/to/raw.csv",
                  col_names = FALSE,
                  col_types = cols(.default = "c"),
                  trim_ws = FALSE)
View(input)
```
<br>

::: {.row}
::: {.box}
[1]{.box-title}
**`col_names = FALSE`**

Headers count as data — row 1 is row 1.
:::

::: {.box}
[2]{.box-title}
**all character**

No silent type coercion. You decide types via `setObsVar(type = ...)`.
:::

::: {.box}
[3]{.box-title}
**rows & cols from 1**

What you see in `View()` is what the schema addresses.
:::
:::

## The four building blocks
::: {.subtitle}
set* functions you compose with the pipe
:::

::: {.col}

::: {.box}
**`setFormat()`** — How to interpret raw values: decimal separator, thousands separator, NA strings, flags.
:::

::: {.box}
**`setFilter()`** — Drop rows or columns before extraction (footnotes, totals, blanks).
:::

::: {.box}
**`setIDVar()`** — Identifying (qualitative) variables — what the data is about (year, region, animal).
:::

::: {.box}
**`setObsVar()`** — Observed (measured) variables — the numbers (heads, hectares, tonnes).
:::

:::

## Schema patterns reference
::: {.subtitle}
the moves you'll see in real tables
:::

::: {.row}
::: {}
*single column*
```r
setIDVar(name = "ADM1", columns = 2)
```

*implicit (same for whole file)*
```r
setIDVar(name = "ADM0", value = "Mongolia")
```

*wide ID — names spread across columns*
```r
setIDVar(name = "animal",
         columns = c(6:14), rows = 1)
```
:::

::: {}
*obs values + unit factor*
```r
setObsVar(name = "heads",
          columns = 5, factor = 1000)
```

*listed obs (key/value layout)*
```r
setObsVar(name = "heads",
          key = 3, value = "bovine")
```

*format + filter quirks*
```r
setFormat(decimal = ",", na_values = c("...","-")) |>
setFilter(rows = .find(pattern = "footnote",
                       col = 1), invert = TRUE)
```
:::
:::

## .find() and step-by-step debugging
`.find()` — dynamic cell location when positions vary across files

```r
.find(pattern = "cattle")                          # any row, any col
.find(fun = is.numeric, col = 1)                   # only column 1
.find(pattern = "total", col = 1, relative = TRUE) # relative to cluster top-left
.find(pattern = "^total$", col = 1, invert = TRUE) # everything except matches
```

Debugging a schema step by step

```r
schema_val <- validateSchema(schema = schema, input = input)   # resolve .find() calls
input_val  <- validateInput(schema = schema_val, input = input) # apply filters

getIDVars(schema = schema_val, input = input_val)      # one element per cluster
getObsVars(schema = schema_val, input = input_val)
getClusterVar(schema = schema_val, input = input_val)

reorganise(input = input, schema = schema)             # full output
```

## schema_builder() — the easy way

::: {.subtitle}
an interactive Shiny app for building schemas
:::

::: {.row}
::: {.box}
[How it works]{.box-title}

1. Load your raw table:
   `schema_builder(input)`

2. Point and click on cells in the table view:
    - mark ID variables
    - mark observed variables
    - mark clusters and filters

3. App generates the schema and previews `reorganise()` output live.

4. Copy the generated schema code into your script.
:::

::: {.box}
[Why use it]{.box-title}

**Faster.** No looking up argument names; the app exposes the right options for what you click.

**Visual.** You see what you've assigned and what's still ambiguous before running anything.

**Same output.** Generates ordinary `set*()` calls. The schema is portable, version-controllable, scriptable.
:::
:::

## Archetype 1 — tidy

::: {.subtitle}
one row per observation; headers in row 1 — schema is almost trivial
:::

::: {.archetype-tag}
tabs2shift\$tidy
:::

::: {.row}
::: {}
**Raw input**

```{r}
#| eval: true
#| echo: false
kable(tabs2shift$tidy, format = "html") |>
  kable_styling(bootstrap_options = "condensed", font_size = 13)
```
:::

::: {}
**Schema**

```r
setIDVar(name = "territories", columns = 1) |>
setIDVar(name = "year",        columns = 2) |>
setIDVar(name = "commodities", columns = 3) |>
setObsVar(name = "harvested",  columns = 5) |>
setObsVar(name = "production", columns = 6,
                               factor  = 0.1)
```
:::
:::

## Archetype 2 — one wide ID

::: {.subtitle}
an ID variable's values masquerade as headers, with observed variables nested below
:::

::: {.archetype-tag}
tabs2shift\$one_wide_id
:::

::: {.row}
::: {}
**Raw input**

```{r}
#| eval: true
#| echo: false
kable(tabs2shift$one_wide_id, format = "html") |>
  kable_styling(bootstrap_options = "condensed", font_size = 12)
```
:::

::: {}
**Schema**

```r
setIDVar(name = "territories", columns = 1) |>
setIDVar(name = "year",        columns = 3) |>
setIDVar(name = "commodities", columns = c(4, 6),
                               rows    = 1) |>
setObsVar(name = "harvested",  columns = c(4, 6),
                               top     = 2) |>
setObsVar(name = "production", columns = c(5, 7),
                               top     = 2)
```
:::
:::

## Archetype 3 — listed observed variable

::: {.subtitle}
obs variable names appear as values of an ID column; values share one column
:::

::: {.archetype-tag}
tabs2shift\$listed_column
:::

::: {.row}
::: {}
**Raw input**

```{r}
#| eval: true
#| echo: false
kable(tabs2shift$listed_column, format = "html") |>
  kable_styling(bootstrap_options = "condensed", font_size = 12)
```
:::

::: {}
**Schema**

```r
setIDVar(name = "territories", columns = 1) |>
setIDVar(name = "year",        columns = 2) |>
setIDVar(name = "commodities", columns = 3) |>
setObsVar(name = "harvested",  columns = 7,
                               key   = 6,
                               value = "harvested") |>
setObsVar(name = "production", columns = 7,
                               key   = 6,
                               value = "production")
```
:::
:::

## Archetype 4 — implicit variable

::: {.subtitle}
one ID variable is in metadata or filename, not in the table itself
:::

::: {.archetype-tag}
tabs2shift\$implicit_variable
:::

::: {.row}
::: {}
**Raw input**

```{r}
#| eval: true
#| echo: false
kable(tabs2shift$implicit_variable, format = "html") |>
  kable_styling(bootstrap_options = "condensed", font_size = 12)
```
:::

::: {}
**Schema**

```r
setCluster(id = "territories",
           left = 1, top = 4) |>
setIDVar(name = "territories",
         value = "unit 1") |>
setIDVar(name = "year",        columns = 4) |>
setIDVar(name = "commodities", columns = 1) |>
setObsVar(name = "harvested",  columns = 2) |>
setObsVar(name = "production", columns = 3)
```
:::
:::

## Archetype 5 — horizontal clusters

::: {.subtitle}
the same block repeats side-by-side; cluster ID is the variable that differs
:::

::: {.archetype-tag}
tabs2shift\$clusters_horizontal
:::

**Raw input**

```{r}
#| eval: true
#| echo: false
kable(tabs2shift$clusters_horizontal, format = "html") |>
  kable_styling(bootstrap_options = "condensed", font_size = 11)
```

**Schema**

```r
setCluster(id   = "territories",
           left = c(1, 6), top = 2) |>
setIDVar(name = "territories", columns = c(1, 6),
                               rows    = 2) |>
setIDVar(name = "year",        columns = c(2, 7)) |>
setIDVar(name = "commodities", columns = c(1, 6)) |>
setObsVar(name = "harvested",  columns = c(3, 8)) |>
setObsVar(name = "production", columns = c(4, 9))
```

# Mongolia

::: {.tagline}
from raw NSO file to tidy long table
:::

**What we'll do, live:**

1. Open the raw NSO livestock CSV with `col_names = FALSE`
2. Launch `schema_builder()` and click through the variables
3. Run `reorganise()` — see tidy output

::: {.heads-up}
We'll keep adding to the script as we work through the remainder of this deck.
:::

# 2. arealDB

::: {.tagline}
structure, harmonisation, integration
:::

## Every asset moves through three stages

::: {.subtitle}
vocabularies, geometries, tables — same lifecycle
:::

::: {.row}
::: {.box .stage}
[stage 1]{.box-title}
**raw original**

whatever the source ships — zip, xlsx, shp, pdf
:::

::: {.arrow}
→
:::

::: {.box .stage}
[stage 2]{.box-title}
**cleaned**

UTF-8 CSV / GeoPackage; uniform format, original layout preserved
:::

::: {.arrow}
→
:::

::: {.box .stage}
[stage 3]{.box-title}
**normalised**

harmonised against gazetteer / ontology; ready to query
:::
:::

## The directory layout is the database
```{.dir-tree}
<root>/
  inventory.rds                 metadata index
  db_info.RData                 version, authors, licence, top level

  vocabularies/
    stage1/                     raw vocabulary archives
    stage2/                     UTF-8 CSV: <vName>__<dSeries>.csv
    stage3/                     <vName>_terms.parquet, <vName>_levels.parquet
    schemas/                    saved tabshiftr schemas
    mappings/                   <vName>_<dSeries>.csv: source_label → canonical_id

  geometries/
    stage1/                     raw geometry archives (zip, gpkg, shp, ...)
    stage2/                     GeoPackages, named by arealDB convention
    stage3/                     one gpkg per top-level unit (per nation)

  tables/
    stage1/<dSeries>/           raw downloaded files, grouped by source
    stage2/                     UTF-8 CSV copies, named by arealDB convention
    stage3/                     one .rds per top-level unit, harmonised
    schemas/                    saved tabshiftr schemas
```


## Two functions per asset: register and normalise

| asset | register | normalise |
|---|---|---|
| **vocabulary** | `regVocabulary()` | `normVocabulary()` |
| **geometry**   | `regGeometry()`   | `normGeometry()`   |
| **table**      | `regTable()`      | `normTable()`      |

<br>

::: {.row}
::: {.box .stage}
[regX()]{.box-title}

Records metadata in `inventory.rds`. Waits for you to place the actual file in stage1 / stage2.
:::

::: {.box .stage}
[normX()]{.box-title}

Reads stage2, harmonises against gazetteer / ontology, writes stage3. Vocabularies are normalised first.
:::
:::

## Vocabularies: backbone vs. mapping

::: {.subtitle}
the first contributor sets the canonical IDs; everyone else maps to them
:::

::: {.row}
::: {.box}
[Backbone]{.box-title}

First dataseries registered against a vocabulary.

Defines canonical IDs and class structure.

`normVocabulary()` writes:
`<vName>_terms.parquet`
`<vName>_levels.parquet`

*Schema can be term-grain (one row per term) or leaf-grain (one row per leaf, ancestors as columns).*
:::

::: {.box}
[Mapping]{.box-title}

Every subsequent contributor.

Source labels are matched against the backbone.

Confirmed mappings persist in:
`vocabularies/mappings/`
`<vName>_<dSeries>.csv`

*Unmatched labels open `match_builder()` Shiny app for interactive resolution.*
:::
:::

## Geometries: spatial boundaries
::: {.subtitle}
one per administrative level
:::

```r
regGeometry(...,                # named territory selector, e.g. ADM0 = "Mongolia"
            gSeries,            # dataseries name (e.g. "gadm")
            label,              # mapping gazetteer class -> column name
                                # list(ADM1 = "NAME_1")
            archive,            # stage1 filename
                                # "archive.zip|inner.gpkg" for compressed sources
            [...],
            overwrite = FALSE)
```
<br>

::: {.row}
::: {.box}
**Place the file**
Raw archive → `geometries/stage1/`
GeoPackage → `geometries/stage2/<filename>`
:::
::: {.box}
**CRS must be EPSG:4326**
`normGeometry()` harmonises every stage2 file against the gazetteer; writes one stage3 GeoPackage per top-level unit. Unmatched names open `match_builder()`.
:::
:::

## Tables: where the schema meets arealDB
```r
regTable(...,                 # named territory selector, e.g. ADM0 = "Mongolia"
         subset = NULL,       # optional label, no underscores; appears in stage2 name
         dSeries,             # dataseries name for the table
         gSeries,             # dataseries name for the geometry to join against
         label,               # gazetteer level: "ADM0", "ADM1", "ADM2"
         begin, end,          # integer years
         schema,              # tabshiftr schema (must exist in global env)
         archive,
         archiveLink,
         downloadDate,
         updateFrequency,
         diagnose = FALSE,    # test-run reorganise() without saving
         overwrite = FALSE)

normTable(pattern   = NULL,
          query     = NULL,        # optional dplyr-style filter
          ontoMatch = NULL)        # column(s) to match against an ontology, e.g. "animal"
```


# 3. From one country to many

::: {.tagline}
our Mongolia script in the context of the full GPW pipeline
:::

## 00_main.R — the orchestrator
::: {.subtitle}
boot · vocabularies · global sources · national sources · tie together
:::

```r
library(arealDB); library(tabshiftr); library(tidyverse); library(lubridate)

db_path <- "./gld-1km/workflow/01-build_census_database/"

# setup and start database
source(paste0(db_path, "01_boot.R"))

# build vocabulary basis
source(paste0(db_path, "02_vocab.R"))

# global / regional dataseries
source(paste0(db_path, "03_fao.R"))
source(paste0(db_path, "03_agriwanet.R"))
source(paste0(db_path, "03_eurostat.R"))

# national dataseries (one file per country)
source(paste0(db_path, "04_argentina.R"))
source(paste0(db_path, "04_australia.R"))
# ... 17 more countries ... and now 04_mongolia.R

# tie everything together
adb_backup()
adb_archive(outPath = db_path, compress = FALSE)
```


## 01_boot.R · 02_vocab.R
::: {.subtitle}
initialise the database, then bootstrap the vocabularies
:::

::: {.row}
::: {}
**01_boot.R — adb_init()**

```r
model_version <- "0.8.3"
authors <- list(
  cre = list("Steffen Ehrmann"),
  aut = list(cens = c("Katya Perez Guzman")),
  ctb = list(cens = c("Ivelina Georgieva"))
)

adb_init(
  root    = paste0(db_path, "_censusDB"),
  version = model_version,
  licence = "https://creativecommons.org/...",
  author  = list(
    cre = authors$cre,
    aut = authors$aut$cens,
    ctb = authors$ctb$cens),
  level   = "ADM0"
)
```
:::

::: {}
**02_vocab.R — UN gazetteer + GPW ontology**

```r
# 1. gazetteer (UN M49 geoscheme)
regDataseries(name = "unsd", ...)

schema_gazetteer_unsd <-
  setIDVar(name = "global",     columns = 2) |>
  setIDVar(name = "region",     columns = 4) |>
  setIDVar(name = "sub_region", columns = 6) |>
  setObsVar(name = "ADM0",      columns = 9,
            type = "character")

regVocabulary(name = "gazetteer",
              dSeries = "unsd",
              schema = schema_gazetteer_unsd,
              ...)
normVocabulary(pattern = "gazetteer__unsd")

# 2. commodity ontology (GPW livestock)
# same pattern with a different schema
```
:::
:::

::: {.footnote .green}
Vocabularies first — geometries and tables match against them.
:::

## Every country file follows the same pattern

::: {.subtitle}
1. dataseries · 2. geometries · 3. tables
:::

::: {.step-card}
::: {.num}
1
:::
::: {.body}
::: {.label}
regDataseries()
:::
::: {.desc}
one call per source — geometry provider, statistics agency, etc.
:::
:::
:::

::: {.step-card}
::: {.num}
2
:::
::: {.body}
::: {.label}
regGeometry() ×N + normGeometry()
:::
::: {.desc}
register one geometry per administrative level (ADM0, ADM1, ADM2), then normalise all of them at once
:::
:::
:::

::: {.step-card}
::: {.num}
3
:::
::: {.body}
::: {.label}
regTable() ×N + normTable()
:::
::: {.desc}
register one table per subset (e.g. one per animal); ontoMatch resolves the concept column against the ontology
:::
:::
:::

## End-to-end checklist

::: {.box}
[Adding a new data source]{.box-title}

1. Find the source, locate the data
2. Place raw files in `stage1/` subfolders
3. Inspect raw tables with `View(read_csv(..., col_names = FALSE))`
4. Write the tabshiftr schema; test with `reorganise()`
5. `regDataseries()` for every source
6. `regVocabulary()` if new terms; then `normVocabulary()`
7. `regGeometry()` — export GeoPackage (EPSG:4326) to `stage2/`
8. `normGeometry()` — resolve unmatched names in `match_builder`
9. `regTable()` with `diagnose = TRUE` first
10. `normTable()` — resolve concepts in `match_builder`
11. Verify: `readRDS(stage3/<TopUnit>.rds)`
:::

::: {.footnote .green}
**Hands-on: pick a country not in the deck and write its `04_*.R`**
:::

## Common failure modes

::: {.box}
[Watch out for]{.box-title}

**`regGeometry` / `regTable` errors**
`regDataseries()` not called first

**Function hangs at readline prompt**
file not yet in stage1 / stage2 — place it, press a key

**`adb_path` not set**
new R session; re-run `adb_init()`

**`reorganise()` returns wrong columns**
schema column numbers don't match `col_names = FALSE` view

**NAs where values expected**
`na_values` in `setFormat()` doesn't cover actual NA strings

**Unmatched territories**
label points to wrong column, or CRS isn't EPSG:4326

**Schema works in dev but `normTable` fails**
stage2 CSV re-saved by Excel — re-export UTF-8
:::
